Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex
improvements in theorems

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a strong ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)=G^+$.
\end{proposition}

\begin{proof}
Since we start from an empty graph, every pair of vertices $X, Y$ where
 $X \in \tetxrm{An}_G(Y)$ is d-separated by the empty set and is
 connected by the edge $X \to Y$. No other edges are added. Therefore, 
the result is $G^+$. 
\end{proof}

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)
\supseteq G^+$.
\end{proposition}

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \tetxrm{An}_G(Y)$, the edge is 
$X \to Y$ just like when using a strong ancestral oracle; since only correctly oriented
edges are added between ancestor-descendant pairs, such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, an arbitrary 
edge is added in a manner that preserves acyclicity. The result is a supergraph
of $G^+$.
\end{proof}

@@

By combining the previous three propositions, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a strong or weak 
ancestral oracle $\mathcal{A}_G$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_G),\mathcal{D}_G)=G.$
\end{theorem}

@@

\begin{proof}
The loop in Algorithm~\ref{algo:expert} terminates if and only if $E=E^*$. If the loop does not 
terminate, then a new edge has been added in line 6, or one or more edges were pruned in line 7.
Every edge can be added at 
most once and pruned at most once. Therefore, the algorithm always terminates
after at most $|V|(|V|-1)+1$ iterations of the loop. For every edge $e=X\to Y $ in the skeleton of $G$, $\mathcal{D}_G(X,Y,\mathbf{Z})=0$ irrespective of $\mathbf{Z}$. Therefore, $e$ must be added to $E$ in some iteration, and can never be pruned afterwards. Therefore, after some iteration, $(V,E)$ must be a supergraph of $G$ after executing line 6, and this will be pruned to the real graph $G$ in line 7 (Proposition~\ref{prop:prune}). In the next iteration, no further changes are made, and the loop terminates.
\end{proof}

>>>>

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a strong ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)=G^+$.
\label{prop:strongexpand}
\end{proposition}

\begin{proof}
Since we start from an empty graph, every pair of vertices $X, Y$ where
 $X \in \textrm{An}_G(Y)$ is d-separated by the empty set and is
 connected by the edge $X \to Y$. No other edges are added. Therefore, 
the result is $G^+$. 
\end{proof}

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)
\supseteq G^+$.
\label{prop:weakexpand}
\end{proposition}

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ just like when using a strong ancestral oracle; since only correctly oriented
edges are added between ancestor-descendant pairs, such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, an arbitrary 
edge is added in a manner that preserves acyclicity. The result is a supergraph
of $G^+$.
\end{proof}

@@

\begin{proof}
For an edge $X \to Y$ in the skeleton of $G$, $\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$ regardless 
of $\mathbf{Z}$, so all edges in the skeleton are retained after {\sc Prune}. Conversely, if 
 $X \to Y$ is not in the skeleton of $G$, let $\mathbf{Z}$ be the $d$-separating set chosen 
in line~5 of {\sc Prune} (there is at least one such $\mathbf{Z}$, the union of the parents of 
$X$ and $Y$). Since  $\mathbf{Z}$ $d$-separates all paths from $X$ to $Y$ in $G'$, it does the same 
in $G$ which contains a subset of these paths. Therefore, all edges not in the skeleton of $G$ are 
removed.
\end{proof}

By combining Propositions~\ref{prop:strongexpand}, \ref{prop:weakexpand} and 
\ref{prop:prune}, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a strong or weak 
ancestral oracle $\mathcal{A}_G$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_G,\emptyset,\infty),\mathcal{D}_G)=G.$
\end{theorem}

@@

\begin{proof}
The loop in Algorithm~\ref{algo:expert} terminates if and only if $E=E^*$. If the loop does not 
terminate, a new edge has been added in line 6, and/or one or more edges were pruned in line 7; the pruned 
edges never include the edge added in line 6 in the same iteration.
Every edge can be added at 
most once and pruned at most once. Therefore, the algorithm always terminates
after at most $|V|(|V|-1)+1$ iterations of the loop. For every edge $e=X\to Y $ in the skeleton of $G$, $\mathcal{D}_G(X,Y,\mathbf{Z})=0$ irrespective of $\mathbf{Z}$, and $e$ must be added to $E$ in some iteration, and can never be pruned afterwards. Therefore, after some iteration, $(V,E)$ must be a supergraph of $G$ after executing line 6, and this will be pruned to the real graph $G$ in line 7 (Proposition~\ref{prop:prune}). In the next iteration, no further changes are made, and the loop terminates.
\end{proof}
